---
title: "Case Study 1"
subtitle: "Doing Data Science 405"
output: md_document
author: "Dave Dyer"
theme: united
---

## Introduction
Data are freely available from the worldbank on both Education levels and initiatives from various countries and also general data on gross domestic product for each country.  When combined and analyzed, one can derive useful insight into how gdp and education might be related.  This is an observational study, so no causal inferences can be drawn, but the relationships between the data sets are interesting and lead to more questions that could theoretically be used to help inform future decisions about education initiatives of the world bank.  In particular, this study looks at the question of whether or not OECD countries have a higher GDP than non-OECD countries.  

## Getting Data
This bit of code pulls down two csv files from the worldbank data sets and downloads it to a subfolder of the project called "data".  Note:  If the directory doesn't yet exist, this will create it.

```{r message=FALSE}
source("cs1_getdata.R") 
```


## Tidying and Merging Data  
The GDP data set has some header rows, some extraneous rows at the end, some columns we're not interested in, some blank lines and some blank data.  

Before merging with the Education data set, this set needs cleaned up.  In short, we remove the extraneous or blank columns, remove blank rows and rows with missing data, and trim off the header and footer.  

Once complete. there are 190 rows with full data in the GDP data set.  Now the data are ready to be merged with the Education data set (which is significantly cleaner.)  The code for loading the data into memory, cleaning the gdp data set and merging with the Education data set is found in cs1_mergedata.R.

```{r message=FALSE}
source("cs1_mergedata.R")
```

After removing the missing or irrelevant rows in the GDP data, there were 190 rows left that show country (both name & code), GDP and ranking.  For the education data, there are 234 rows.  The difference between the two data sets can be found using the setdiff() set operation:

```{r message=FALSE, echo=FALSE, results='hide'}
missing <- setdiff(edu$CountryCode, gdp$CountryCode)
for (code in missing) {
  print(edu$Long.Name[edu$CountryCode==code])
}
```


The reason these are missing from the merged data is because there was no data collected from them in the world bank gdp data.

The following code merges the data into a single data set.  Note that rows with NA's or empty fields within the education data set were not yet removed, since not all analysis requires a complete data set in all columns.

Now that the getting, cleaning and merging of the data are all complete, we can move on to answering interesting questions about the data set.



### Question 1:
Merging on the field "CountryCode", there are 189 rows that match (there were 190 'complete' rows in the GDP data.)  The reason for the mis match between 234 total rows and 189 matching rows is explained above using the setdiff operation.  

The one economy that is missing from the education data set, but appears in the gdp data set is "South Sudan".  The GDP data has both South Sudan and Sudan, whilst the education data only has "Republic of Sudan."  The time from 2010 to present has been a volatile time for Sudan, with confusing national borders, economies, widespread violence, elections, etc.  For the purposes of this study, we will consider Sudan (SDN) GDP, since there is a match in both data sets, and not South Sudan GDP, which we will consider an 'incomplete' case.


### Question 2:
After sort in ascending order by GDP, using the following code:

```{r echo=FALSE}
all_sorted <- all_the_datas[order(all_the_datas$GDP2012, decreasing=FALSE),]
print(all_sorted$Long.Name[13])
```
The Country with the 13th lowest GDP is St. Kitts and Nevis.

### Question 3:
The organization for economic cooperation and development (OECD) was established to promote policies that would improve the economic and social well-being of people around the world (http://www.oecd.org/about/).  An interesting question to analyze is that of whether or not OECD countries with high Income have higher-ranked GDPs on average than Non-OECD, yet still high-income countries.

To calculate the average GDP rankings for the High Income: OECD and High Income: nonOECD groups, we use the following code:

```{r echo=FALSE}
all <- all_sorted # tired of typing underscores and it doesn't matter if it's sorted or not when taking averages of rank
highOECD <- all[all$Income.Group=="High income: OECD",]
highNonOECD <- all[all$Income.Group=="High income: nonOECD",]
avg_high_oecd <- mean(as.numeric(highOECD$Rank))
avg_high_non_oecd <- mean(as.numeric(highNonOECD$Rank))
avg_high_oecd
avg_high_non_oecd
```

The average rank of high income OECD countries is 32.97, while the average rank of non-oecd high income countries is 91.91, almost 60 ranks lower on average.  While this is just an observational study, and we therefore cannot attributed the rank difference to OECD membership, it is nonetheless interesting and worth further exploration.

### Question 4:
Below is a plot of the GDP for all countries.  Note that the GDP is plotted on a log scale so that you can see the countries with the smaller GDPs.  Also, the bars are colored by their Income.Group.  

Note also that there are some missing lables in the plot.  Those are (from top to bottom, respectively):

1.  Rep\\xfablica Bolivariana de Venezuela 
2.  Democratic Republic of S\\xe3o Tom\\xe9 and Principe
3.  Republic of C\\xf4te d'Ivoire 

There is clearly an issue with encoding characters in the data file.  If I have time I will go back and manually fix this.

```{r fig.width=10, fig.height=28}
library(ggplot2)
data <- all

ggplot(all, aes(x=Long.Name, y=GDP2012)) + 
  geom_bar(stat='identity', aes(fill=Income.Group), width=0.75) +
  labs(title = "GDP (in Millions) by Country", subtitle="Color = Income Group;  GDP is Plotted on Log Axis") +
  coord_flip() + 
#  theme(text= element_text(size=06)) + 
  scale_y_log10(label=scales::comma)
```

### Question 5:
To illustrate how GDP Ranking compares with Income group, we cut the GDP rankings into 5 quantile groups and compare against income groups in a table.  

To get the quantile cutoffs from the rank, we run the quantile() function.  
```{r}
rank_quantile <- quantile(as.numeric(all$Rank), probs = c(0,20,40, 60, 80, 100)/100.0)
ranks_quantile_vector <- c(rank_quantile[[1]], rank_quantile[[2]], rank_quantile[[3]], rank_quantile[[4]], rank_quantile[[5]], rank_quantile[[6]])
```

We set the ranking percentage cutoffs at 0%, 20%, 40%, 60%, 80%, and 100%, with the following results for gdp rank quantiles.

* 0% - 2
* 20% - 40.6
* 40% - 78.2
* 60% - 114.8
* 80% - 152.4
* 100% - 190.0

As expected, the ranks are about evenly split amongst the 5 groups since since the ranks are in order from 1 to 190 with equal steps between each rank, and no breaks.

When put into a table to compare Income.Group with Rank Quantile, we see that 4 of the low income group are ranked in the top 20% of all countries' GDP, while 4 of the High Income OECD countries are ranked in the lower 20% of all countries' GDP.
```{r echo=FALSE}
all$qrank <- cut(as.numeric(all$Rank), 
                 breaks=quantile(as.numeric(all$Rank), c(0,0.2, 0.4, 0.6, 0.8, 1.0)), 
                 include.lowest=TRUE)
ranktable <- table(all$Income.Group, all$qrank)
ranktable
```

Of Lower Middle Income countries, there are 13 within the top 38 nations with the highest GDP.

## Conclusion
In general, one would think that OECD membership would place countries higher on the GDP-ranking scale, but based on our preliminary analysis, that may not be the case.  Of the top 38 ranked countries, 68% are from Upper and Lower Middle income groups, and only 7% are from High Income OECD groups.  


